Chemistry of the Orphan Stream I: Identifying Stream Candidates with Low Resolution Spectroscopy

### 2012-09-12 2012-oss-mrs/Response_GDC_20120912_II.txt

Ok, now the draft.  Given my previous message I think 4.3 is likely to get revised a bit along the lines:
- discuss Ca II metallicities (with the calibration issues, etc)
- discuss isochrone metallicities given Newburg distance.
- correlate these estimates
==> sample of most probable members  (and less likely, and probably not, members)
Then given the probable members, likely first revise distance estimate - will give minor adjustment to isochrone abundances.
Show isochrone fit.  Then in probably new section discuss:
a) metallicity spread of probable members - noting if significant, then expect progenitor likely dSph.  If small, then progenitor more likely a GC - as in Wylie de Boer et al 2012 ApJ 755 35 for Aquarius.  Probably can't say too much because will depend on confirmation of membership, e.g. the metal-poor star.  But we can speculate.
b) the line of site velocity dispersion - seemingly very small.

More specific comments.  (Happy with responses to my earlier comments).

Abstract.  I think technically the AAT is still the Anglo-Australian Telescope despite the name change of the AAO.  May be stick with AAT?

>> You're right. I knew that one of them (AAO/AAT) had changed names and the other had not. I got them mixed up. Corrected in text.

Second last sentence of Abstract.  I'd prefer something like "The observed velocity dispersion of our highest probability members is consistent with arising from the velocity errors alone, indicating that, at least as regards the
line-of-sight, the Orphan Stream is kinematically quite cold.

>> That's better: Changed.

top of right hand column page 1, first 4 lines:
"total low luminosity" ==> "low total luminosity"
>> Changed.

"in reliably identifying" ==> "to separate reliably"
>> Altered to "to reliably separate Orphan Stream members from halo stars": "to separate reliably" didn't quite make sense in context.

"will be detected" ==> "will be established"
>> Changed.

and I think the Keller et al reference should go after "SkyMapper" rather than at the end of the sentence, although not a big issue.
>> Placed after "SkyMapper telescope".

Next para - maybe "Consequently, the fainter…"
>> Changed

maybe an "and" between Xue et al and Koposov et al rather than a semi-colon.
>> Added

maybe "kinematic maps of these streams"  - we're not talking specifically of the OS here.
>> Changed to "More complete photometric and kinematic maps of these low total luminosity streams may provide the best test as to whether this mass discrepancy is an artefact of incomplete observations."

Last para:
"seven highly probable Orphan Stream candidates"  - the 'seven' might need changing.  It is after all 5 stars in section 5.
>> #TODO

page 2, left column 3rd line.  Surely one of the discovery papers can fill "[citation needed]".
>> I checked them all just now, and none of them make this claim. I can't remember where I first got this number from, but it's about right. I don't think it's too big of a claim to leave there without a citation. I've removed the [citation needed], and I'm happy to see what the referee says.

3 lines from the bottom: "discriminate dwarfs from giants"  (i.e. and ==> from)
>> Changed

2 lines from the bottom "spectra" is missing after "arm".
>> Added

2nd paragraph of 4.1.  Think we need to say that "stars" in the Besancon models were selected in the same way as the observed sample, i.e. same colour and magnitude cuts.  (New Fig 1 is fine).
>> Whoops! I had forgotten to do this after Stefan gave me the true selection criteria. I've done it now though and re-plotted the histogram. The kinematic distribution did not change significantly. I've also added the following as the second sentence in 2nd paragraph of 4.1: " We have selected predicted stars from Besan\c{c}on using the same criterion outlined in \S\ref{sec:target-selection} after using the \citet{Jordi;et-al_2006} colour transformations."

5th line of 4.2 on page 3 "narrow" ==> "narrower"
>> Changed

(Haven't made any specific comments on current 4.3 - not appropriate if it's going to be revised as I suggested).

2nd para of Section 5.  I think the words "brighter than our current limit of g ~ 18" need to go in after "members" as it doesn't quite make sense otherwise.  
>> Sorry: I can't find this anywhere in the text.

3rd para - probably put "i.e., consistent with zero" after "0.3 +/- 3.3 kms-1".  Again the numbers might change if the sample of members changes.  And this actual paragraph will probably now come before the conclusions section.

OK, enough!

Cheers,
Gary

### 2012-09-12 2012-oss-mrs/draft-oss-mrs.tex
Incorporated all comments from Comments_GDC_20120912_II.txt

\begin{abstract}
We present coordinates and properties of candidate K giant members in the Orphan Stream, which have been identified from low resolution spectroscopic data taken with AAOmega on the Australian Astronomical Telescope. With modest S/N and independent cuts in photometry, kinematics, gravity and metallicity we yield self-consistent, highly probable stream members. Our data indicates an overall stream metallicity of [Fe/H] $= -1.78 \pm 0.08$ which is more metal-rich than previously found and unbiased by spectral type. We find a revised stream distance of 24 kpc near the celestial equator, and our kinematic signature peaks at $V_{helio} \sim 218$\,km s$^{-1}$. The Orphan Stream appears extremely cold, with an intrinsic dispersion of less than $0.3 \pm 3.3$ km s$^{-1}$. We highlight likely members for high resolution spectroscopic follow-up.
\end{abstract}

@@

The Orphan Stream was independently detected by both \citet{Grillmair;Dionatos_2006} and \citet{Belokurov;et-al_2006}, and is distinct from other substructures in the halo. The stream stretches over $60\,^\circ$ in the sky, has a low surface brightness and a narrow stream width of only $\sim2^\circ$ and as the name suggests, the parent object largely remains a mystery. The stream extends past the celestial equator \--- outside the SDSS footprint \--- but is not detected in existing southern surveys \citep{Newberg;et-al_2010}. Whilst the parent system remains elusive, significant effort has been placed on associating the stream with known Milky Way satellites \citep{Zucker;et-al_2006, Fellhaur;et-al_2007,Jin;Lynden_Bell_2007,Sales;et-al_2008}. In contrast, there has been relatively limited observational work on the Orphan Stream itself other than the original discovery papers \citep{Grillmair;Dionatos_2006, Belokurov;et-al_2006, Belokurov;et-al_2007} and the work of \citet{Newberg;et-al_2010}. This is largely to be expected given the absence of multi-band photometry in the southern sky and the total low luminosity of the stream, making it difficult in reliably identifying Orphan Stream members from halo stars. It is unlikely the full extent of the Orphan Stream will be detected until the SkyMapper telescope surveys this region \citep{Keller;et-al_2007}.

As \citet{Sales;et-al_2008} points out, there is a natural observational bias towards more massive and recent mergers like Sagittarius. The fainter end of this substructure distribution has yet to be fully recovered, or thoroughly examined. Interestingly, there are indications that  some fainter substructures like the Orphan Stream \citep{Newberg;et-al_2010} and the Palomar 5 tidal tails \citep{Odenkirchen;et-al_2009} have orbits which seem to be best-fit by Milky Way models with nearly 60\% less mass \citep{Newberg;et-al_2010} than generally reported by \citet{Xue;et-al_2008, Koposov;et-al_2010}. Such a discrepancy in the mass of the Milky Way is worrying. A more complete photometric and kinematic map of the stream may provide the best test as to whether this mass discrepancy for faint substructure orbits is real or an artefact of incomplete observations. Whilst the full spatial extent of the stream remains incomplete, we can examine the detailed chemistry of the stream and investigate the stream history, as well as make predictions about the nature of the progenitor.

@@

We have targeted K giant members of the Orphan Stream in order to investigate their detailed chemistry. Because K-giants are difficult to unambiguously detect from photometry alone, low resolution spectroscopy is required to estimate stellar parameters and determine radial velocities. The Orphan Stream has a low spatial over-density on the order of $\sim10\%$ [citation needed], which makes it difficult to separate stream members from halo stars. However, there is a well described distance gradient along the stream \citep{Belokurov;et-al_2007, Newberg;et-al_2010} which provides an indication on where we should focus our spectroscopic efforts.

@@

Observations took place on the Australian Astronomical Telescope using the AAOmega spectrograph in April 2009. AAOmega is a fibre-fed, dual beam multi-object spectrograph which is capable of simultaneously observing spectra of 400 (science and sky) targets across a $2^\circ$ field of view. We used the 5700\,{\AA} dichroic in combination with the 1000I grating in the red arm, and the 580V grating in the blue arm. This provides a spectral coverage between $800 \leq \lambda \leq 950$\,nm in the red at $\mathcal{R} \approx 4400$, and between $370 \leq \lambda \leq 580$\,nm with a lower resolving power of $\mathcal{R} \approx 1300$ in the blue.

The data were reduced using the standard \textsc{2DFDR} reduction pipeline \footnote{http://www.aao.gov.au/2df/aaomega/aaomega\_2dfdr.html}. After flat-fielding, throughput calibration for each fibre was achieved using the intensity of skylines in each fibre. The median flux of dedicated sky fibres was taken for sky subtraction, and wavelength calibration was performed using arc lamp exposures between science frames. Three thirty minute science exposures were median combined to assist with cosmic ray removal. The median S/N obtained in the red arm for our fields is modest at 35\,px$^{-1}$. With the presence of strong Ca \textsc{II} triplet lines in the red arm we are able to ascertain reliable radial velocities and reasonable estimates on overall metallicity \citep[][and references therein]{Starkenburg;et-al_2010}. Our spectral region also includes gravity-sensitive magnesium lines: Mg I at 8807\,{\AA}, and the Mg I\,b 3$p$-4$s$ triplet lines at $\sim$5178\,{\AA}. As we demonstrate in the next section, these lines are sufficient to discriminate dwarfs and giants even with weak signal.

The blue and red arm were normalised using a third order cubic spline after multiple iterations of outlier clipping. We used defined knot spacings of 20 nm in the red arm, and 5 nm in the blue arm in order to accommodate often poor S/N, and varying strengths of molecular band-heads.

@@

% Besan\c{c}on heliocentric velocities have been translated to a galactocentric frame in the same way as our targets.
Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin;et-al_2003}. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{\textsc{gsr}} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg;et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. No statistically significant peak representative of the Orphan Stream exists in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

We have measured the equivalent widths of the gravity-sensitive Mg I line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g \-- r$) and metallicity, giant stars present narrow Mg I absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In four cases no Mg I 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS $g \-- r$ de-reddened\footnote{All magnitudes presented in this paper are de-reddened}, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

Although OSS 6 has an uncertain radial velocity, using a maximum-likelihood estimation we find the mean stream galactocentric velocity at $(l, b) = (250\,^\circ, 50\,^\circ)$ from five stars to be $V_{GSR} = 82 \pm 1.6$ km s$^{-1}$ and the intrinsic dispersion to be $0.3 \pm 3.3$ km s$^{-1}$. For our four stars with low radial velocity errors this dispersion is found to be $0.00 \pm 2.4$ km s$^{-1}$. These velocities indicate that the Orphan Stream is a very cold stream, and implies their motion is largely perpendicular to the line of sight.

>>>>

\begin{abstract}
We present coordinates and properties of candidate K giant members in the Orphan Stream, which have been identified from low resolution spectroscopic data taken with AAOmega on the Anglo-Astronomical Telescope. With modest S/N and independent cuts in photometry, kinematics, gravity and metallicity we yield self-consistent, highly probable stream members. Our data indicates an overall stream metallicity of [Fe/H] $= -1.78 \pm 0.08$ which is more metal-rich than previously found and unbiased by spectral type. We find a revised stream distance of 24 kpc near the celestial equator, and our kinematic signature peaks at $V_{helio} \sim 218$\,km s$^{-1}$. The observed velocity dispersion of our highest probability members is consistent with arising from the velocity errors alone, indicating that, at least as regards the
line-of-sight, the Orphan Stream is kinematically quite cold. We highlight likely members for high resolution spectroscopic follow-up.
\end{abstract}

@@

The Orphan Stream was independently detected by both \citet{Grillmair;Dionatos_2006} and \citet{Belokurov;et-al_2006}, and is distinct from other substructures in the halo. The stream stretches over $60\,^\circ$ in the sky, has a low surface brightness and a narrow stream width of only $\sim2^\circ$ and as the name suggests, the parent object largely remains a mystery. The stream extends past the celestial equator \--- outside the SDSS footprint \--- but is not detected in existing southern surveys \citep{Newberg;et-al_2010}. Whilst the parent system remains elusive, significant effort has been placed on associating the stream with known Milky Way satellites \citep{Zucker;et-al_2006, Fellhaur;et-al_2007,Jin;Lynden_Bell_2007,Sales;et-al_2008}. In contrast, there has been relatively limited observational work on the Orphan Stream itself other than the original discovery papers \citep{Grillmair;Dionatos_2006, Belokurov;et-al_2006, Belokurov;et-al_2007} and the work of \citet{Newberg;et-al_2010}. This is largely to be expected given the absence of multi-band photometry in the southern sky and the low total luminosity of the stream, making it difficult to reliably seperate Orphan Stream members from halo stars. It is unlikely the full extent of the Orphan Stream will be established until the SkyMapper telescope \citep{Keller;et-al_2007} surveys this region.

As \citet{Sales;et-al_2008} points out, there is a natural observational bias towards more massive and recent mergers like Sagittarius. Consequently, the fainter end of this substructure distribution has yet to be fully recovered, or thoroughly examined. Interestingly, there are indications that  some fainter substructures like the Orphan Stream \citep{Newberg;et-al_2010} and the Palomar 5 tidal tails \citep{Odenkirchen;et-al_2009} have orbits which seem to be best-fit by Milky Way models with nearly 60\% less mass \citep{Newberg;et-al_2010} than generally reported by \citet{Xue;et-al_2008} and \citet{Koposov;et-al_2010}. Such a discrepancy in the mass of the Milky Way is worrying. More complete photometric and kinematic maps of these low total luminosity streams may provide the best test as to whether this mass discrepancy is an artefact of incomplete observations. Whilst the full spatial extent of the stream remains incomplete, we can examine the detailed chemistry of the stream and investigate the stream history, as well as make predictions about the nature of the progenitor.

@@

We have targeted K giant members of the Orphan Stream in order to investigate their detailed chemistry. Because K-giants are difficult to unambiguously detect from photometry alone, low resolution spectroscopy is required to estimate stellar parameters and determine radial velocities. The Orphan Stream has a low spatial over-density on the order of $\sim10\%$, which makes it difficult to separate stream members from halo stars. However, there is a well described distance gradient along the stream \citep{Belokurov;et-al_2007, Newberg;et-al_2010} which provides an indication on where we should focus our spectroscopic efforts.

@@

Observations took place on the Anglo-Astronomical Telescope using the AAOmega spectrograph in April 2009. AAOmega is a fibre-fed, dual beam multi-object spectrograph which is capable of simultaneously observing spectra of 400 (science and sky) targets across a $2^\circ$ field of view. We used the 5700\,{\AA} dichroic in combination with the 1000I grating in the red arm, and the 580V grating in the blue arm. This provides a spectral coverage between $800 \leq \lambda \leq 950$\,nm in the red at $\mathcal{R} \approx 4400$, and between $370 \leq \lambda \leq 580$\,nm with a lower resolving power of $\mathcal{R} \approx 1300$ in the blue.

The data were reduced using the standard \textsc{2DFDR} reduction pipeline \footnote{http://www.aao.gov.au/2df/aaomega/aaomega\_2dfdr.html}. After flat-fielding, throughput calibration for each fibre was achieved using the intensity of skylines in each fibre. The median flux of dedicated sky fibres was taken for sky subtraction, and wavelength calibration was performed using arc lamp exposures between science frames. Three thirty minute science exposures were median combined to assist with cosmic ray removal. The median S/N obtained in the red arm for our fields is modest at 35\,px$^{-1}$. With the presence of strong Ca \textsc{II} triplet lines in the red arm we are able to ascertain reliable radial velocities and reasonable estimates on overall metallicity \citep[][and references therein]{Starkenburg;et-al_2010}. Our spectral region also includes gravity-sensitive magnesium lines: Mg I at 8807\,{\AA}, and the Mg I\,b 3$p$-4$s$ triplet lines at $\sim$5178\,{\AA}. As we demonstrate in the next section, these lines are sufficient to discriminate dwarfs from giants even with weak signal.

The blue and red arm spectra were normalised using a third order cubic spline after multiple iterations of outlier clipping. We used defined knot spacings of 20 nm in the red arm, and 5 nm in the blue arm in order to accommodate often poor S/N, and varying strengths of molecular band-heads.

@@

% Besan\c{c}on heliocentric velocities have been translated to a galactocentric frame in the same way as our targets.
Figure \ref{fig:velocities} shows a histogram of our galactocentric velocities, compared to the predicted smooth line-of-sight velocity distribution for this region from the Besan\c{c}on model \citep{Robin;et-al_2003}. We have selected predicted stars from Besan\c{c}on using the same criterion outlined in \S\ref{sec:target-selection} after using the \citet{Jordi;et-al_2006} colour transformations. It is clear that our target selection has yielded mostly nearby thick disk dwarf stars with $V_{\textsc{gsr}} \approx -120$\,km s$^{-1}$. The expected kinematic signature for the Orphan Stream, as \citet{Newberg;et-al_2010} reported, is labelled in Figure \ref{fig:velocities}. No statistically significant peak representative of the Orphan Stream exists in our sample. From kinematics alone, our targets appears largely indistinguishable from a smooth halo distribution.

@@

We have measured the equivalent widths of the gravity-sensitive Mg I line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g \-- r$) and metallicity, giant stars present narrower Mg I absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In four cases no Mg I 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS $g \-- r$ de-reddened\footnote{All magnitudes presented in this paper are de-reddened}, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

@@

Although OSS 6 has an uncertain radial velocity, using a maximum-likelihood estimation we find the mean stream galactocentric velocity at $(l, b) = (250\,^\circ, 50\,^\circ)$ from five stars to be $V_{GSR} = 82 \pm 1.6$ km s$^{-1}$ and the intrinsic dispersion to be $0.3 \pm 3.3$ km s$^{-1}$, i.e., consistent with zero. For our four stars with low radial velocity errors this dispersion is found to be $0.00 \pm 2.4$ km s$^{-1}$. These velocities indicate that the Orphan Stream is a very cold stream, and implies their motion is largely perpendicular to the line of sight.
